\begin{center} \bf \sffamily \LARGE Prisoner's dilemma with stochastic payoffs\\[-1cm] \end{center}

### 2021-03-25 main.tex
parts of writing

\documentclass[11pt]{article}
\usepackage{times}
\usepackage{amsmath,amsthm,amssymb,setspace,enumitem,epsfig,titlesec,verbatim,color,array,eurosym,multirow}
\usepackage[sort&compress]{natbib}
\usepackage[footnotesize,bf]{caption}
\usepackage[margin=2.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{standalone}
\usepackage{tikz}
\smallskip % Erlaubt kleine Abstaende zwischen Paragraphen, falls es dem Seitenlayout hilft
\renewcommand{\baselinestretch}{1.3}
\newcommand{\R}{\mathbb{R}}

@@

Most existing evolutionary game studies work with idealized assumptions. As
stated in~\cite{Roca2006} on of these assumptions is that individuals play many
times and with all other players before reproduction takes place, so that
payoffs, equivalently fitness, are given by the mean distribution of types in
the population. Both hypotheses, common in biological evolution, imply that
selection occurs much more slowly than the interaction between individuals.
Although recent experimental studies show that this may not always be the case
in biology, it is clear that in cultural evolution or social learning the time
scale of selection is much closer to the time scale of interaction. And these
assumptions can change the outcome of the dynamics. The work of~\cite{Roca2006}
we show that rapid selection affects evolutionary dynamics in such a dramatic
way that for some games it even changes the stability of equilibria.

When modeling how individuals make decisions in each round, these models assume
that players only remember the last round. However, when modeling how
individuals update their strategies over time, individuals are assumed to have
perfect memory. Here, we explore how robust our understanding of cooperation is
as models deviate from the perfect memory assumption~\cite{Traulsen2007,
Baek2016, Ye2017}.

Evolutionary game dynamics have extensively been studied from mostly
deterministic models based on rate equations to stochastic individual-based
models. These more sophisticated models
These more sophisticated models use a microscopic process as a starting
point to determine how successful strategies spread. As stated in %ToDo references
there are two classes have been used extensively: (i)
fitness-based processes in which an individual chosen proportional to fitness
reproduces and the offspring replaces a randomly chosen individual%ToDo references
Pairwise comparison processes in which a pair of individuals is chosen, and
where subsequently one of these individuals may adopt the strategy of the other.
This adoption occurs with a probability that depends on the payoff of both
individuals, such that better players are more likely to be imitated than those
who do worse. %ToDo references

We follow a Pairwise comparison processes. $N$ individuals interact by playing a
game and reproduce by selecting one individual, with probability proportional to
the payoff, to duplicate and substitute a randomly chosen individual. The payoff
of every player is set to zero after each reproduction event, and this two-step
cycle is repeated until the population eventually stabilizes. This stochastic
dynamics is discrete in both population and time, while keeping the population
size constant over time. Interestingly, this microscopic dynamics leads to a
difference equation that has been proposed as an adjusted [1] or discrete-time
[2] analogous of the replicator equation, widely used in evolutionary game
theory (see [9] for a recent, detailed discussion of this issue). Additionally,
we note that for social applications, reproduction may be also interpreted as a
learning process, in which individuals do not die but instead change the way
they behave or their strategies.

As for the game, we consider the donation game where each player can cooperate by providing 
a benefit \(b\) to the other player at their cost \(c\), with \(0 < c < b\).

@@

However, we will also consider the important case of symmetric \(2 \times 2\) games,
in which the payoffs are given by the following matrix:

\begin{equation}
\begin{pmatrix}
    R & S  \\
    T & P
\end{pmatrix}
\end{equation}

For the donation game (\(T=b, R=b-c, S=-c, P=0\), and matrix~(\ref{eq:pd_normal_form})).

Let n be the number of individuals using strategy 1, also referred as type 1
individuals. After each reproduction event n may stay the same, increase by one,
or decrease by one.

The individual chosen for reproduction A replaces B with probability p, which
depends on the payoff difference between the two individuals. The composition of
the population can only change if both individuals are of different types. We
follow (Blume, 1993, Szabó and Tőke, 1998, Hauert and Szabó, 2005) in choosing
the Fermi function from statistical physics for p

n both types of processes, the relative influence of the game is controlled by
an external parameter, the so- called intensity of selection $\beta$. This parameter
has strong parallels to the inverse temperature in statistical mechanics [16]

\section{R.1}

>>>>

\documentclass[11pt]{article}
\usepackage{times}
\usepackage{amsmath,amsthm,amssymb,setspace,enumitem,epsfig,titlesec,verbatim,color,array,eurosym,multirow}
\usepackage[sort&compress]{natbib}
\usepackage[footnotesize,bf]{caption}
\usepackage[margin=2.5cm, includefoot, footskip=30pt]{geometry}
\usepackage{standalone}
\usepackage{tikz}
\usepackage{subcaption}
\smallskip % Erlaubt kleine Abstaende zwischen Paragraphen, falls es dem Seitenlayout hilft
\renewcommand{\baselinestretch}{1.3}
\newcommand{\R}{\mathbb{R}}

@@

Traditionally, evolutionary game studies work with idealized assumptions when
calculating fitness. As stated in~\cite{Roca2006}, it is commonly assumed that
individuals play many times and with all other players before reproduction takes
place, so that payoffs, equivalently fitness, are given by the mean distribution
of types in the population. Initially, this implies that selection occurs much
more slowly than the interaction between individuals, even experimental studies
show that this may not always be the case in biology. Based on %ToDo references
it is clear that in cultural evolution or social learning the time scale of
selection is much closer to the time scale of interaction. The work
of~\cite{Roca2006} relaxed this assumption and showed that rapid selection
affects evolutionary dynamics in such a dramatic way that for some games it even
changes the stability of equilibria.

Furthermore, the second assumption implies that individuals have the time to
interact with with all other players. Moreover, this implies that individuals
when update their strategies over time, individuals are assumed to have perfect
memory. However, when modeling how individuals make decisions in each round,
these models assume that players only remember the last
round~\cite{Traulsen2007, Baek2016, Ye2017}. The aim of this work is too relax
this assumption, and to further the work of~\cite{Roca2006} by also considering
strategies that interact to the past interactions.

In literature, when individuals update their strategies over time they interact
with all other players so that the are given by the mean distribution of types
in the population. These payoffs are referred to as the \textit{expected}
payoffs of individuals. We contrast this standard scenario with an alternative
scenario where individuals update their strategies when remembering several of
their past interactions but not all of them. We refer to these as the \textit{stochastic}
payoffs of individuals.

Evolutionary game dynamics have extensively been studied from mostly
deterministic models based on rate equations to stochastic individual-based
models. These more sophisticated models make use of different rules to determine
how successful strategies spread. As stated in~\cite{Wu2015} there are two
classes have been used extensively, \textit{fitness-based processes} in which an
individual chosen proportional to fitness reproduces and the offspring replaces
a randomly chosen individual and \textit{pairwise comparison} processes in which
a pair of individuals is chosen, and where subsequently one of these individuals
may adopt the strategy of the other. This adoption occurs with a probability
that depends on the payoff of both individuals, such that better players are
more likely to be imitated than those who do worse.

In this we work we follow a pairwise comparison processes. We consider a well
mixed population of $N$ individuals. In our model mutation happens rarely and
such each individual can be of one of the two types, either of the
\textit{resident} or either of the \textit{mutant}.

Regarding the game, we the important case of symmetric \(2 \times 2\) games,
in which the payoffs are given by the following matrix:

\begin{equation}\label{eq:game}
    \begin{pmatrix}
        R & S  \\
        T & P
    \end{pmatrix}
\end{equation}

These cases include the Snowdrift~\cite{sugden2004economics}, Stag Hunt~\cite{skyrms2001stag},
Harmony~\cite{Licht1999} and the Prisoner's Dilemma games. These results are 
presented in Section~\ref{section:2_by_2_games}.

Moreover, we will specifically consider the donation game where each player can
cooperate by providing a benefit \(b\) to the other player at their cost \(c\),
with \(0 < c < b\). Then, \(T=b, R=b-c, S=-c, P=0\), and matrix~(\ref{eq:games})
is give by:

@@

These results are presented in Section~\ref{section:donation}. The relative
influence of the game is controlled by an external parameter, the so called
intensity of selection $\beta$.

The space of strategies in repeated is infinite. To continue with our
evolutionary analysis, we assume herein that individuals at most make use of
simple \textit{reactive strategies}. Reactive strategy are a set of memory-one
strategies that only take into account the previous action of the opponent. An
example of a reactive strategy is Tit For Tat. Reactive strategies can be
written explicitly as a vector \(\in \R_{3}\). More specifically a reactive
strategy \(s\) is given by \(s=(y, p, q)\) where \(y\) is the probability that
the strategy opens with a cooperation and \(p, q\) are the probabilities that
the strategy cooperates given that the opponent cooperated and defected
equivalently. The individuals of our population do not interact for a given
number of turns, on the contrast, we consider that each match has a probability
$\delta$ to continue following each turn.

\section{Expected and stochastic payoffs in the donation game}\label{section:donation}

Figure~\ref{fig:expected_and_stochastic_for_donation} shows simulation results
for the above described process. Figure~\ref{fig:expected_and_stochastic_for_donation}~depicts the evolving conditional
cooperation probabilities $p$ and $q$ (assuming that the discount
factor~$\delta$ and the benefit $b$ are comparably high). The left panel
corresponds to the standard scenario considered in the previous literature. It
considers players who use expected payoffs to update their
strategies. The right panel shows the scenario considered herein, in which
players update their strategies based on their last round's payoff. The figure
suggests that when updating is based on expected payoffs, players tend to be
more generous (their $q$-values are higher on average). In addition, players are
generally more cooperative.

\begin{figure}[!htbp]
    \includegraphics[width=\textwidth]{static/expected_and_stochastic_for_donation_game.pdf}
    \caption{{\bf Evolutionary dynamics under expected payoffs and stochastic payoffs.} 
    We have run two simulations of the evolutionary process described in
    Section~\ref{section:methods} for $T\!=\!10^7$ time steps. For each time step,
    we have recorded the current resident population ($y,p,q$). Since simulations
    are run for a relatively high continuation probability of $\delta\!=\!0.999$, we
    do not report the players' initial cooperation probability $y$. The graphs show
    how often the resident population chooses each combination ($p,q$) of
    conditional cooperation probabilities in the subsequent rounds. ({\bf A}) If
    players update based on their expected payoffs, the resident population
    typically applies a strategy for which $p\!\approx\!1$ and
    $q\!\le\!1\!-\!c/b\!=\!0.9$. The cooperation rate within the resident population
    (averaged over all games and over all time steps) is close to 100\%. ({\bf B})
    When players update their strategies based on their realized payoffs in the last
    round, there are two different predominant behaviors. The resident population
    either consists of defectors (with $p\!\approx\!q\!\approx\!0$) or of
    conditional cooperators. In the latter case, the maximum level of $q$ consistent
    with stable cooperation is somewhat smaller compared to the expected-payoff
    setting, $q\!<\!0.5$. Also the resulting cooperation rate is smaller. On
    average, players cooperate roughly in half of all rounds. Parameters:
    $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
    \label{fig:expected_and_stochastic_for_donation}
\end{figure}

The difference in the cooperation rates becomes more and more obvious as the
temptation to defect increases, Figure~\ref{fig:expected_and_stochastic_for_donation_b_10}.

\begin{figure}[!htbp]
    \includegraphics[width=\textwidth]{static/expected_and_stochastic_for_donation_game_b_10.pdf}
    \caption{{\bf Evolutionary dynamics under expected payoffs and stochastic payoffs.} 
    Parameters:
    $N\!=\!100$, $b\!=\!10$, $c\!=\!1$, $\beta\!=\!1$, $\delta\!=\!0.999$.}
    \label{fig:expected_and_stochastic_for_donation_b_10}
\end{figure}

We explored how the evolving cooperation rates change as we vary the benefit $b$
and the selection strength $\beta$
(Figure~\ref{fig:cooration_under_parameters}). In all cases, we find that the
two scenarios yield similar cooperation rates when the respective parameters are
small. Once there is a high benefit to cooperation, strong selection, or a high
expected number of rounds, updating based on expected payoffs yields higher
cooperation rates.

\begin{figure}
    \centering
    \begin{subfigure}{.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{static/cooperation_rate_over_b.pdf}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
      \centering
      \includegraphics[width=\textwidth]{static/cooperation_rate_over_betas.pdf}
    \end{subfigure}
\caption{{\bf The evolution of cooperation for different parameter values.} 
While the previous figures depict the evolutionary outcome for fixed parameter
values, here we vary the benefit of cooperation $b$ and the strength of
selection $\beta$. In all cases, stochastic payoff evaluation tends to reduce
the evolving cooperation rates. Unless explicitly varied, the parameters of the
simulation are $N\!=\!100$, $b\!=\!3$, $c\!=\!1$, $\beta\!=\!1$,
$\delta\!=\!0.99$. Simulations are run for $T\!=\!5\times 10^6$ time steps for
each parameter combination.}\label{fig:cooration_under_parameters}
\end{figure}

\section{Expected and stochastic payoffs in $2 \times 2$ games}\label{section:2_by_2_games}

\section{Methods}\label{section:methods}

In evolution context we consider a population of \(N\) players,
where \(N\) is even, and where mutations are sufficiently rare. At any point
in time the there are at most two different strategies are present in
the population. Suppose there are \(N - k\) players who use the strategy \(s_{1}=(y_{1},
p_{1}, q_{1})\), whereas \(k\) players use the strategy \(s_{2}=(y_{2}, p_{2},
q_{2})\). We refer to these two player types as ‘residents’ and ‘mutants’,
respectively.

Each step of the evolutionary process consists of two stages, a game stage and
an updating stage.

\begin{enumerate}
    \item In the game stage, each player is randomly matched with some
    other player in the population to interact in one instance of the IPD.
    \item In the updating stage, two players are randomly drawn from
    the population, a `learner' and a `exemplar'. Given that the learner's payoff
    in the last round is $u_L\!\in\! \mathcal{U}$ and that the exemplar's last
    round's payoff $u_E\!\in\! \mathcal{U}$, we assume the learner adopts the role
    model's strategy with probability

    \begin{equation} \label{Eq:rho}
    \rho(u_L, u_E) = \frac{1}{1\!+\!\exp\big[ \!-\!\beta (u_E\!-\!u_L) \big]}. 
    \end{equation}

    where $\beta\!\ge\!0$ corresponds to the strength of selection.
\end{enumerate}

We iterate this basic evolutionary step until either the mutant strategy goes
extinct, or until it fixes in the population (in which case the mutant strategy
becomes the new resident strategy). After either outcome, we introduce a new
mutant strategy $s'_2\!=\!(y_2',p_2',q_2')$ (uniformly chosen from all reactive
strategies at random), and we set the number of mutants to $k\!=\!1$. This
process of mutation and fixation/extinction is then iterated many times.

We compare this process for what we defined as {\bf stochastic payoff
evaluation} with the analogous process where players update their strategies
with respect to their {\bf expected} payoffs,

\begin{equation} \label{eq:ExpPay}
\begin{array}{lcrcr}
\pi_1 &= &\displaystyle \frac{N\!-\!k\!-\!1}{N-1}\cdot \langle\mathbf{v}(s_1,s_1),\mathbf{U}\rangle	& + &\displaystyle\frac{k}{N-1}\cdot \langle\mathbf{v}(s_1,s_2),\mathbf{U}\rangle,\\[0.5cm]
\pi_2 &= &\displaystyle\frac{N-k}{N-1}\cdot \langle\mathbf{v}(s_2,s_1),\mathbf{U}\rangle & + &\displaystyle\frac{k-1}{N-1}\cdot \langle\mathbf{v}(s_2,s_2),\mathbf{U}\rangle.\\
\end{array}
\end{equation}

In the limit of no discounting, $\delta\!\rightarrow\! 1$, this process based on
expected payoffs has been considered in~\cite{imhof2010stochastic}.

We define {\bf stochastic payoff} as the average payoff $u\!\in\! \mathcal{U}$ a player
receives in the last \(n\) rounds of the game given that they interact with
\(m\) players.

\subsubsection*{Case \(n=m=1\).}

Initially, consider the situation where \(n=m=1\). The player's stochastic
payoff is what they receive in the last round against a single opponent. There
only four possible outcomes for the last round, those are \(CC, CD, DC, DD\).
Consider two players with reactive strategies $S_1\!=\!(y_1, p_1, q_1$) and
$S_2\!=\!(y_2,p_2,q_2)$ who interact in a repeated prisoner's dilemma with
continuation probability $\delta$, the probability that are in each of the
four possible states in the last round is given by:

\begin{equation}
    \mathbf{v}(s_1,s_2)\!=\!\Big(\mathbf{v}_{R}(s_1,s_2),\mathbf{v}_{S}(s_1,s_2),\mathbf{v}_{T}(s_1,s_2),\mathbf{v}_{P}(s_1,s_2)\Big).
\end{equation}

where,

\begin{equation} \label{Eq:LastRound}
    \setlength{\arraycolsep}{1pt}
    \begin{array}{rcl}

    \mathbf{v}_{R}(S_1,S_2) &= &\displaystyle (1\!-\!\delta)\frac{y_1y_2}{1\!-\!\delta^2 r_1 r_2}+\delta \frac{\Big(q_1+r_1\big((1\!-\!\delta)y_2+\delta q_2\big)\Big) \Big(q_2+r_2\big((1\!-\!\delta)y_1+\delta q_1\big)\Big)}
    {\displaystyle(1\!-\!\delta r_1r_2)(1\!-\!\delta^2 r_1 r_2)},\\[1cm]

    \mathbf{v}_{S}(S_1,S_2) &= &\displaystyle (1\!-\!\delta)\frac{y_1\bar{y}_2}{1\!-\!\delta^2 r_1 r_2}+\delta \frac{\Big(q_1+r_1\big((1\!-\!\delta)y_2+\delta q_2\big)\Big) \Big(\bar{q}_2+\bar{r}_2\big((1\!-\!\delta)y_1+\delta p_1\big)\Big)}
    {\displaystyle(1\!-\!\delta r_1r_2)(1\!-\!\delta^2 r_1 r_2)},\\[1cm]

    \mathbf{v}_{T}(S_1,S_2) &= &\displaystyle (1\!-\!\delta)\frac{\bar{y}_1y_2}{1\!-\!\delta^2 r_1 r_2}+\delta \frac{\Big(\bar{q}_1+\bar{r}_1\big((1\!-\!\delta)y_2+\delta p_2\big)\Big) \Big(q_2+r_2\big((1\!-\!\delta)y_1+\delta q_1\big)\Big)}
    {\displaystyle(1\!-\!\delta r_1r_2)(1\!-\!\delta^2 r_1 r_2)},\\[1cm]

    \mathbf{v}_{P}(S_1,S_2) &= &\displaystyle (1\!-\!\delta)\frac{\bar{y}_1\bar{y}_2}{1\!-\!\delta^2 r_1 r_2}+\delta \frac{\Big(\bar{q}_1+\bar{r}_1\big((1\!-\!\delta)y_2+\delta p_2\big)\Big) \Big(\bar{q}_2+\bar{r}_2\big((1\!-\!\delta)y_1+\delta p_1\big)\Big)}
    {\displaystyle(1\!-\!\delta r_1r_2)(1\!-\!\delta^2 r_1 r_2)}.
    \end{array}
\end{equation}

\begin{proof}

Assume a repeated prisoner's dilemma between two reactive strategies. Given the
continuation probability $\delta$, probability that the game ends in the after
the first round $(1 - \delta)$ and the expected distribution of the four
outcomes in the very first round is $\mathbf{v_0}$ defined as. Following the
first round the, the outcome of the next rounds with a probability $\delta$
is \(M\) such that,

\[\dots\]

It can shown that, \(!(1\!-\!\delta)\mathbf{v_0}(I_4-\delta M)^{-1}\) and with
some algebraic manipulation we derive to Equation~\ref{Eq:LastRound}.

\end{proof}
